---
title: "housing_model"
author: "Sarai Saenz & Sofia Hernandez"
date: "2026-04-15"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE)
```

## Setup

This is where we will add all our dependencies.

```{r}

packages <- c("caret", "ggplot2", "corrplot", "dplyr", "VIM", "e1071")
installed <- packages %in% rownames(installed.packages())
if (any(!installed)) install.packages(packages[!installed])

library(caret)
library(ggplot2)
library(corrplot)
library(dplyr)
library(VIM)
library(e1071)
```

## -------- Load and Inspect Data -------- 

The Housing data set contains many missing values and categorical variables. Our main task involves imputing or dropping missing values, encoding categoricals, and log transforming the target variable SalePrice since it is right skewed.

```{r}
train <- read.csv("train.csv", stringsAsFactors = FALSE)
test <- read.csv("test.csv", stringsAsFactors = FALSE)

cat("Train dimensions:", dim(train), "\n")
cat("Test dimensions:", dim(test), "\n")

# Save the target (Sales Price) and IDs before merging
train_ids <- train$Id
test_ids <- test$Id
sale_price <- train$SalePrice

#Remove ID and target from train set, ID from test
train$Id <- NULL
test$Id <- NULL
train$SalePrice <- NULL
```

### Merging Train and Test
Merging allows us to ensure any factor levels or imputed medians are computed on the full data distribution, all train and test columns should match.

```{r}
# Tag each set so we can come split them later
train$set <- "train"
test$set <- "test"
all_data <- bind_rows(train, test)

cat("Combined dimensions", dim(all_data), "\n")

```

### Inspect Missing Values
We have some variables such as GarageType that have observations marked as NA. However, that just means there is no garage, not that there is a missing distinction. This means we have got to check the none_cols and zero_cols list against the data description file.

```{r}
missing_counts <- sort(colSums(is.na(all_data)),decreasing = TRUE)

missing_pct <- missing_counts / nrow(all_data) * 100

missing_df <- data.frame(
  variable = names(missing_counts),
  count = missing_counts,
  percentage = round(missing_pct, 1)
)

cat("\nTop 40 variables with missing values:\n")
print(head(missing_df[missing_df$count > 0, ], 40))
```

So we can see that PoolQC and Misc Feature are missing in almost every row, this isn't an error. That just means that 99% of those houses don't have those items. The columns referring to Garage all have exactly 159 missing values, meaning that 159 houses don't have garages at all. This is similar to the Basement group however there is a slight variation 80 vs 82 that suggests that a rating might have been forgotten. 

### Fix NAs that mean "None"
Now that we have inspected our None values, we can remove the true NA's and save the "None" values.

```{r}
none_cols <- c(
  "Alley", "BsmtQual", "BsmtCond", "BsmtExposure", "BsmtFinType1",
  "BsmtFinType2", "FireplaceQu", "GarageType", "GarageFinish",
  "GarageQual", "GarageCond", "PoolQC", "Fence", "MiscFeature",
  "MasVnrType"
)
for (col in none_cols) {
  if (col %in% names(all_data)) {
    all_data[[col]][is.na(all_data[[col]])] <- "None"
  }
}
 
# Numeric "None" equivalents (e.g. no basement = 0 sq ft)
zero_cols <- c(
  "BsmtFinSF1", "BsmtFinSF2", "BsmtUnfSF", "TotalBsmtSF",
  "BsmtFullBath", "BsmtHalfBath", "GarageYrBlt", "GarageArea",
  "GarageCars", "MasVnrArea"
)
for (col in zero_cols) {
  if (col %in% names(all_data)) {
    all_data[[col]][is.na(all_data[[col]])] <- 0
  }
}

# Impute Random Categorical Data with the Mode
cat_cols <- names(all_data)[sapply(all_data, is.character)]
for (col in cat_cols) {
  if (any(is.na(all_data[[col]]))) {
    mode_val <- names(sort(table(all_data[[col]]), decreasing = TRUE))[1]
    all_data[[col]][is.na(all_data[[col]])] <- mode_val
  }
}
 
# Numerics: fill with median
num_cols <- names(all_data)[sapply(all_data, is.numeric)]
num_cols <- setdiff(num_cols, "set")
for (col in num_cols) {
  if (any(is.na(all_data[[col]]))) {
    all_data[[col]][is.na(all_data[[col]])] <- median(all_data[[col]], na.rm = TRUE)
  }
}
 
cat("\nMissing values remaining:", sum(is.na(all_data[, names(all_data) != "set"])), "\n")
```

### Feature Engineering

```{r}

# Temp Data Frame for EDA
train_analysis <- all_data[all_data$set == "train", ]
train_analysis$SalePrice <- sale_price

# Identify strongest predictors
numeric_vars <- sapply(train_analysis, is.numeric)
cor_matrix <- cor(train_analysis[, numeric_vars], use = "pairwise.complete.obs")

price_correlations <- sort(cor_matrix[,"SalePrice"], decreasing = TRUE)
print(head(price_correlations, 15))
```

```{r}
ggplot(train_analysis, aes(x = as.factor(OverallQual), y = SalePrice)) +
  geom_boxplot(fill = "steelblue") +
  labs(title = "Sale Price vs Overall Quality")

ggplot(train_analysis, aes(x = reorder(Neighborhood, SalePrice, FUN = median), y = SalePrice)) +
  geom_boxplot() +
  coord_flip() +
  labs(title = "Sale Price by Neighborhood (Ordered by Median)")
```

```{r}
# --------- Feature Engineering ----------

# Total square footage
all_data$TotalSF <- all_data$TotalBsmtSF + all_data$X1stFlrSF + all_data$X2ndFlrSF
 
# Total bathrooms
all_data$TotalBaths <- all_data$FullBath + (0.5 * all_data$HalfBath) +
                       all_data$BsmtFullBath + (0.5 * all_data$BsmtHalfBath)
 
# Age
all_data$HouseAge   <- all_data$YrSold - all_data$YearBuilt
all_data$RemodelAge <- all_data$YrSold - all_data$YearRemodAdd
 
# Has features flags
all_data$HasPool     <- as.integer(all_data$PoolArea > 0)
all_data$HasGarage   <- as.integer(all_data$GarageArea > 0)
all_data$HasFireplace<- as.integer(all_data$Fireplaces > 0)
all_data$HasBasement <- as.integer(all_data$TotalBsmtSF > 0)

# Try Squaring OverallQual for polynomial feature
all_data$OverallQual_Squared <- all_data$OverallQual^2
```

